01_Vegetation_cover

### 2022-12-28 Manuscript.Rmd
add to intro

---
output:
  word_document:
    reference_docx: "Manuscript-Markdown-template.docx"
    fig_caption: yes
---

@@

@@

Alternative to forming a direct representation of past plant trait composition, pollen based trait reconstruction might be better suited to study relative changes in trait composition. For instance, @Sande202 showed that pollen assemblages reflect relative trait changes in trait composition of the vegetation.

@@

### Study sites

We collected pollen, trait and species composition data from 16 sites in the Cairngorms National Park, Scotland, United Kingdom. Sites were selected to represent four vegetation types in the area: grassland, heath land, deciduous forest and coniferous forest. Sampling locations were at least 1 km from each other to avoid pollen source areas from overlapping with each other. Pollen, trait and species data was also available from `r nrow(bdm_meta)` sites across Switzerland were monitored in the context of the Biodiversity Monitoring Switzerland (BDM). This is a systematic sampling grid for the long-term monitoring of plant and animal species in Switzerland [@BDM_2022]. Trait and species composition data, but not pollen data, from an additional `r nrow(read.csv("Data/Swiss_LC/all_BDM_plots_coord.csv")) - nrow(bdm_meta)` sites that were surveyed in this network were also used in this study.

The data that was collected in Scotland was used to test if pollen-based trait reconstructions represents vegetation trait composition. We compare the community weighted mean (CWM) trait value of the vegetation with the CWM trait values reconstructed from surface pollen samples. Surface pollen samples are collected from moss polsters and approximate vegetation composition in recent years[@Pardoe2010]. Study design mostly followed the protocol of @Bunting_2013, which allows for the calculation of vegetation composition in concentric circles around the moss sampling point to test the spatial scale that is represented by the moss sample (Appendix). We calculated the vegetation composition in three nested concentric rings of 10 m, 100 m and 1000 m in radius. The inner ring (10 m) and the middle ring was surveyed in the field. The vegetation cover of the outer ring (1000 m) was estimated using QGIS and land cover maps [\@scotisch_government; \@QGIS_software]. To test the effect of spatial scale on the representativeness of pollen assemblages for trait composition reconstruction, we test compare vegetation CWM and pollen CWM in the rings of 10 m, 100 m and 1000 m. Furthermore, four different treatments were assigned to test what factors could influence representation of trait composition by pollen data. Firstly, the use of correction factors, than the separation of vegetation and pollen data in to plant functional type and pollination mode. Lastly, the vegetation was aggregated to genus level and family level to test the effect of taxonomic resolution on the reconstructions.

@@

### Vegetation survey

At the Scottish sites, in the inner ring (10 m) 17 sub-plots of 0.25 m^2^ were surveyed: one in the centre, 3 in every wind direction at 3, 6 and 10 meters, and one in the NE, NW, SE and SW direction (See design in Appendix). In the middle ring (10-100 m) between 4 and 6 plots of 5 by 5 meter were selected on a stratified random design based on vegetation types that were identified in the fields. Vegetation abundance was estimated using percentage cover. In Switzerland, the plots were 1.8 m in radius and contained 4 sub-plots. Each sub-plot was a quadrant of 0.16 m^2^ with nine points (a 3 x 3 grid), and percentage coverage of the species present at the nine points was estimated. All species names were standardized according to the Leipzig Plant Catalogue using the *lcvplants* and *lcvp* R packages [@freiberg2020lcvp].

### Trait data collection

We used three traits for the analysis, specific leaf area, leaf area and and plant height. These traits reflect different components in the spectrum of plant strategies [@Diaz2016]. In Scotland, trait data was collected in the middle ring (100 m). For determining plant height, we randomly selected 10 flowering individuals of each species at every site. Non-flowering individuals were selected if flowering individuals were not available. For determining specific leaf area (SLA) and leaf area (LA), we collected five leaf samples per species in sealable plastic bags. Fresh weight and surface area were measured within 72 hours of collection using a carat scale and flatbed scanner. Samples were dried at 70 ***°***C for 72 hours to constant dry weight before measuring dry weight. In Switzerland, trait data was collected for the species present at the nine points in the 3 by 3 grid, otherwise trait measurement was the same.

For reconstructing trait composition from pollen data and previously gapfilled trait data from the TRY database was used [@kattge2020try] [\@Schrodt_2015]. for classifying vegetation data in to plant growth forms and pollination mode,

### Pollen data

Moss was collected from multiple location within a 1 m radius in the centre of each plot. Only living tissue was sampled. Pollen were extracted from the moss using standardized pollen preparation protocol (REF). Pollen were identified using light microscope and the key of (REF). To test the ability of pollen correction factors to improve pollen - vegetation relationships, pollen productivity estimates from @Githumbi_2022 were used. Pollen productivity estimates were not available for all taxa, taxa without pollen productivity factors were excluded. On average, this led to the exclusion of `r round(mean(missing_ppe_scot$percent_missing),2)*100`% of the Scottish pollen data (max `r round(max(missing_ppe_scot$percent_missing),2)*100`%). Pollen were divided in to two plant functional groups according to @Reitalu_2019 consisting of woody (trees and shrubs) or non-woody species (grasses, herbs and dwarf shrubs of the Ericaceae family). @Reitalu_2019 was also used to classify pollen in to pollination mode, either insect-pollinated or not insect-pollinated.

For reconstructing plant trait composition from pollen data, assumptions needed to be made about the species belonging to the pollen taxa in the data set. Therefore we created a pollen to species translation table. We used the *rgbif* package to download the current distribution of species belonging to each pollen taxon in Scotland and Switzerland from the Global Biodiversity Information Facility database (REF). Species distribution data was queried using the pollen taxonomic name, i.e. the lowest taxonomic name the pollen could be identified to. All trait data that was available for the species in the translation table was then used in the calculation trait values on the pollen taxonomic level .

@@

Where $\alpha$ and $\beta$ are the intercept and the slope, respectively, which are supplied by uninformative priors. $Y_{elev}$ is the elevation as the linear predictor. We modelled this relationship for the grassland and forest plots separately.

@@

Over all treatments and spatial scales, we only found evidence of a significant positive relationship between pollen-based reconstructed CWM trait values and vegetation CWM values for plant height. We found little support for a positive association between pollen and vegetation trait value for leaf area and specific leaf area. The spatial scale matters for the relationship between pollen plant height and vegetation plant height, only at scales between 10-100 m the relationship is significant.

>>>>

---
output:
  word_document:
    reference_docx: "Manuscript-Markdown-template.docx"
    fig_caption: yes
    bibliography: "Validation_study_references.bib" 
---

@@

Because of the difference in  pollen productivity and dispersal among pollen taxa, pollen counts might not reflect absolute abundance and hence trait composition [@Sande2021]. Alternative to forming a direct representation of past plant trait composition, pollen based trait reconstruction might be better suited to study relative changes in trait composition. For instance, @Sande2021 showed that pollen assemblages reflect relative trait changes in trait composition of the vegetation along an elevation gradient.

@@

@@

### Study sites
We collected pollen, trait and species composition data from 16 sites in the Cairngorms National Park, Scotland, United Kingdom. We selected sites to represent four vegetation types in the area: grassland, heath land, deciduous forest and coniferous forest. Sampling locations were at least 1 km from each other to avoid pollen source areas from overlapping with each other. Pollen, trait and species data was also available from `r nrow(bdm_meta)` sites across Switzerland were monitored in the context of the Biodiversity Monitoring Switzerland (BDM). This is a systematic sampling grid for the long-term monitoring of plant and animal species in Switzerland [@BDM_2022]. Trait and species composition data, but not pollen data, from an additional `r nrow(read.csv("Data/Swiss_LC/all_BDM_plots_coord.csv")) - nrow(bdm_meta)` sites that were surveyed in this network were also used in this study.

### Study design
We used the Scottish data to test if pollen-based trait reconstructions represents vegetation trait composition. We compare the community weighted mean (CWM) trait value of the vegetation with the CWM trait values reconstructed from surface pollen samples. Study design mostly followed the protocol of @Bunting_2013, which allows for the calculation of vegetation composition in concentric circles around the moss sampling point to test the spatial scale that is represented by the moss sample (Appendix). We calculated the vegetation composition in three nested concentric rings of 10 m, 100 m and 1000 m in radius. The inner ring (10 m) and the middle ring was surveyed in the field. The vegetation cover of the outer ring (1000 m) was estimated using QGIS and land cover maps [\@scotisch_government; \@QGIS_software]. To test the effect of spatial scale on the representativeness of pollen assemblages for trait composition reconstruction, we compare vegetation CWM and pollen CWM in the rings of 10 m, 100 m and 1000 m. Furthermore, four different treatments were assigned to test what factors could influence representation of trait composition by pollen data. Firstly, the use of correction factors, than the separation of vegetation and pollen data in to plant functional type and pollination mode. Lastly, the vegetation was aggregated to genus level and family level to test the effect of taxonomic resolution on the reconstructions.

@@

### Vegetation survey
At the Scottish sites, we surveyed 17 sub-plots of 0.25 m^2^ in the inner ring (10 m): one in the center, 3 in every wind direction at 3, 6 and 10 meters, and one in the NE, NW, SE and SW direction (See design in Appendix). In the middle ring (10-100 m), we selected between 4 and 6 plots of 5 by 5 meter based on a stratified random design based on vegetation types that were identified in the fields. Vegetation abundance was estimated using percentage cover. In Switzerland, the plots were 1.8 m in radius and contained 4 sub-plots. Each sub-plot was a quadrant of 0.16 m^2^ with nine points (a 3 x 3 grid), and percentage coverage of the species present at the nine points was estimated. All species names were standardized according to the Leipzig Plant Catalogue using the *lcvplants* and *lcvp* R packages [@freiberg2020lcvp].

### Trait data collection
We used three traits for the analysis, specific leaf area, leaf area and and plant height. These traits reflect different components in the spectrum of plant strategies [@Diaz2016]. In Scotland, we collected trait data in the middle ring (100 m). For determining plant height, we randomly selected 10 flowering individuals of each species at every site. Non-flowering individuals were selected if flowering individuals were not available. For determining specific leaf area (SLA) and leaf area (LA), we collected five leaf samples per species in sealable plastic bags. Fresh weight and surface area were measured within 72 hours of collection using a carat scale and flatbed scanner. Samples were dried at 70 ***°***C for 72 hours to constant dry weight before measuring dry weight. In Switzerland, trait data was collected for the species present at the nine points in the 3 by 3 grid, otherwise trait measurement was the same.For reconstructing trait composition from pollen data and previously gapfilled trait data from the TRY database was used [@kattge2020try] [@Schrodt_2015].

### Pollen data
Surface pollen samples are collected from moss polsters and approximate vegetation composition in recent years [@Pardoe2010]. Moss was collected from multiple location within a 1 m radius in the center of each plot.Pollen were extracted from the moss using standardized pollen preparation protocol (REF). Pollen were identified using light microscope and the key of (REF).

### 
To test the ability of pollen correction factors to improve pollen - vegetation relationships, pollen productivity estimates from @Githumbi_2022 were used.Pollen productivity estimates were not available for all taxa, taxa without pollen productivity factors were excluded. On average, this led to the exclusion of `r round(mean(missing_ppe_scot$percent_missing),2)*100`% of the Scottish pollen data (max `r round(max(missing_ppe_scot$percent_missing),2)*100`%). Pollen were divided in to two plant functional groups according to @Reitalu_2019 consisting of woody (trees and shrubs) or non-woody species (grasses, herbs and dwarf shrubs of the Ericaceae family). @Reitalu_2019 was also used to classify pollen in to pollination mode, either insect-pollinated or not insect-pollinated.

For reconstructing plant trait composition from pollen data, assumptions needed to be made about the species belonging to the pollen taxa in the data set. Therefore we created a pollen to species translation table. We used the *rgbif* package to download the current distribution of species belonging to each pollen taxon in Scotland and Switzerland from the Global Biodiversity Information Facility database (REF). Species distribution data was queried using the pollen taxonomic name, i.e. the lowest taxonomic name the pollen could be identified to. All trait data that was available for the species in the translation table was then used in the calculation trait values on the pollen taxonomic level. For reconstructing trait composition from pollen data and previously gapfilled trait data from the TRY database was used [@kattge2020try] [@Schrodt_2015].

@@

Where $\alpha$ and $\beta$ are the intercept and the slope, respectively, which are supplied by uninformative priors. $Y_{elev}$ is the elevation as the linear predictor. We modeled this relationship for the grassland and forest plots separately.

@@

### Pollen representation by the vegetation
- missing species in pollen 
- 
### Effect of spatial scale and plant functional group
Over all treatments and spatial scales, we only found evidence of a significant positive relationship between pollen-based reconstructed CWM trait values and vegetation CWM values for plant height. We found little support for a positive association between pollen and vegetation trait value for leaf area and specific leaf area. The spatial scale matters for the relationship between pollen plant height and vegetation plant height, only at scales between 10-100 m the relationship is significant.
